---
title: "Hands-on Exercise 1"
editor: visual
---

## Overview

This is the overview paragraph.

## Getting Started

In this hands-on exercise, the following packages will be used:

-   **readr** for importing csv data,

-   **readxl** for importing Excel worksheet,

-   **tidyr** for manipulating data,

-   **dplyr** for transforming data, and

-   **ggplot2** for visualising data

```{r}
packages = c('sf', 'tidyverse')

for (p in packages){

if(!require(p, character.only = T)){

install.packages(p)

}

library(p,character.only = T)

}
```

### Importing polygon feature data in shapefile format

```{r}
mpsz = st_read(dsn = "data/geospatial",

layer = "MP14_SUBZONE_WEB_PL")
```

### Importing polyline feature data in shapefile form

```{r}
cyclingpath = st_read(dsn = "data/geospatial",

layer = "CyclingPath")
```

### Importing GIS data in kml format

```{r}
preschool = st_read("data/geospatial/pre-schools-location-kml.kml")
```

### Checking the Content of A Simple Feature Data Frame

#### Working with *st_geometry()*

```{r}
st_geometry(mpsz)
```

#### 
Working with glimpse()

```{r}
glimpse(mpsz)
```

#### Working with *head()*

```{r}
head(mpsz, n=5)
```

### 
Plotting the Geospatial Data

```{r}
plot(mpsz)
```

```{r}
plot(st_geometry(mpsz))
```

```{r}
plot(mpsz["PLN_AREA_N"])
```

### Working with Projection

#### Assigning EPSG code to a simple feature data frame

```{r}
st_crs(mpsz)
```

```{r}
mpsz3414 <- st_set_crs(mpsz, 3414)
```

```{r}
st_crs(mpsz3414)
```

#### Transforming the projection of preschool from wgs84 to svy21.

```{r}
preschool3414 <- st_transform(preschool,

crs = 3414)
```

```{r}
st_geometry(preschool3414)
```

### Importing and Converting An Aspatial Data

#### Importing the aspatial data

```{r}
listings <- read_csv("data/aspatial/listings.csv")
```

```{r}
list(listings)
```

```{r}
listings_sf <- st_as_sf(listings,

coords = c("longitude", "latitude"),

crs=4326) %>%

st_transform(crs = 3414)
```

```{r}
glimpse(listings_sf)
```

```{r}
buffer_cycling <- st_buffer(cyclingpath,

dist=5, nQuadSegs = 30)
```

```{r}
buffer_cycling$AREA <- st_area(buffer_cycling)
```

```{r}
sum(buffer_cycling$AREA)
```

```{r}
mpsz3414$`PreSch Count`<- lengths(st_intersects(mpsz3414, preschool3414))
```

```{r}
summary(mpsz3414$`PreSch Count`)
```

```{r}
top_n(mpsz3414, 1, `PreSch Count`)
```

```{r}
mpsz3414$Area <- mpsz3414 %>%

st_area()
```

```{r}
mpsz3414 <- mpsz3414 %>%

mutate(`PreSch Density` = `PreSch Count`/Area * 1000000)
```

```{r}
hist(mpsz3414$`PreSch Density`)
```

#### #histogram

```{r}
ggplot(data=mpsz3414,

aes(x= as.numeric(`PreSch Density`)))+

geom_histogram(bins=20,

color="black",

fill="light blue") +

labs(title = "Are pre-school even distributed in Singapore?",

subtitle= "There are many planning sub-zones with a single pre-school, on the other hand, \nthere are two planning sub-zones with at least 20 pre-schools",

x = "Pre-school density (per km sq)",

y = "Frequency")
```

#### #scatter plot

```{r}
ggplot(data=mpsz3414,

aes(x= as.numeric(`PreSch Density`), y= as.numeric(`PreSch Count`)))+

geom_point() +

labs(title = "Are pre-school even distributed in Singapore?",

subtitle= "There are many planning sub-zones with a single pre-school, on the other hand, \nthere are two planning sub-zones with at least 20 pre-schools",

x = "Pre-school density (per km sq)",

y = "Pre-school Count")
```

```{r}

```

#### 

#### 
